---
title: "Colon Cancer Survival Analysis"
author: "Sourena Mohit"
output: html_notebook
---

# Step 1: Load and Import the Dataset
In this step, we want to:

1. Ensure the survival package is installed (because it contains the colon dataset).
2. Load the survival package.
3. Load the colon dataset into our R environment.
4. Copy the dataset to a new object (colon_data) so we don’t accidentally overwrite the original.

```{r load-data, message=FALSE, warning=FALSE}

# 1. Check if the 'survival' package is installed; if not, install it.
if (!require("survival")) install.packages("survival", dependencies=TRUE)

# 2. Load the 'survival' package into the R session
library(survival)

# 3. Load the 'colon' dataset from the 'survival' package
data("colon")

# 4. Make a copy of the dataset
colon_data <- colon

```

## Step 2: Explore the Dataset
In Step 2, our goal is to examine the dataset to understand its contents, identify potential issues, and plan how to handle them. Below is a structured explanation of each part of your code and the corresponding results.

```{r}
# Define a helper function to install packages if they're not already installed
install_if_missing <- function(packages) {
  new_pkgs <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(new_pkgs)) install.packages(new_pkgs, dependencies = TRUE)
}

# Specify required packages
packages_needed <- c("skimr", "kableExtra", "dplyr")

# Install missing packages
install_if_missing(packages_needed)

# Load the packages
lapply(packages_needed, library, character.only = TRUE)

# View the first few rows
head(colon_data)

# Check the structure of the dataset
str(colon_data)

# Get summary statistics
summary(colon_data)

# Generate summary statistics using skimr
summary_stats <- skim(colon)

# View the summary statistics
print(summary_stats)

# Convert the summary statistics to a nicely formatted table
summary_table <- summary_stats %>%
  kable("html", caption = "Summary Statistics of the Colon Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# Print the table
summary_table

# Save the table as an HTML file
save_kable(summary_table, file = "summary_table.html")

# Function to plot boxplots for all numeric columns
plot_boxplots <- function(data) {
  numeric_cols <- sapply(data, is.numeric)
  numeric_data <- data[, numeric_cols, drop = FALSE]
  
  # Optional: arrange multiple plots in a grid (2 rows, 3 columns)
  par(mfrow = c(2, 3))  
  for (col in names(numeric_data)) {
    boxplot(
      numeric_data[[col]], 
      main = paste("Boxplot of", col), 
      xlab = col, 
      col = "lightblue", 
      border = "darkblue"
    )
  }
  # Reset the plotting area to a single panel
  par(mfrow = c(1, 1))  
}

# Call the function
plot_boxplots(colon_data)
# Check for missing values
colSums(is.na(colon_data))
```

## Analysis of the Exploratory Results
Below is a structured analysis of your Step 2 exploratory findings. walk through the key takeaways from the summary statistics, missing values check, and boxplots.

### 1. Data Structure Recap
- **Observations**: 1,858 rows (each row can represent a patient event record).
- **Variables**: 16 columns (numeric and factor types).

#### Key Variables:
- **time**: Time in days until an event (recurrence or death) or censoring.
- **status**: Event occurrence indicator (1 = event, 0 = censored).
- **etype**: Event type (1 = recurrence, 2 = death).
- **rx**: Treatment group (3 levels: “Obs”, “Lev”, “Lev+5FU”).
- **nodes**: Number of positive lymph nodes (0–33).
- **differ**: Tumor differentiation (1 = well, 2 = moderate, 3 = poor).
- **sex**: Encoded as 0 (female) and 1 (male), though numeric in the dataset.
- **id**: Patient identifier, which is not directly useful in modeling.

Understanding each variable’s purpose helps in deciding how to handle them during modeling (e.g., which to include as covariates).

### 2. Missing Values
- **nodes**: 36 missing values.
- **differ**: 46 missing values.
- All other variables appear complete.

Missing data can bias results if not handled properly.

### 3. Numeric Variable Distributions and Outliers
From the boxplots:
- **age**: Ranges from 18 to 85; a few younger outliers (below 30).
- **nodes**: Heavily right-skewed. Some patients have a high count (above 10), and the maximum is 33.
- **time**: Spans 8 to 3329 days (over 9 years). The median is around 1855 days (~5 years).
- **sex, obstruct, perfor, adhere, status, surg, node4, etype**: Each is effectively binary or categorical (0/1 or 1/2), so boxplots show minimal variation.
- **differ (1 to 3) and extent (1 to 4)**: Appear mostly around 2–3, with a few outliers at 1 and 4.

Right-skewed data (like nodes and time) may benefit from transformations (e.g., log-transform) if needed for certain models. Outliers might represent genuine clinical scenarios, boxplots confirm the presence of two distinct categories.

### 4. Factor Variables
- **rx**: Three treatment groups (“Obs”, “Lev”, “Lev+5FU”).
- **study**: Always 1 (no variation, so it’s not useful for analysis).

rx is already a factor, but sex might be converted to a factor to make it more intuitive (0 = Female, 1 = Male). Factor variables are crucial in survival models (like Cox regression) to compare hazard rates among groups. Unvarying variables (like study) do not contribute any information to the model.

### 5. Overall Observations
- **No Large-Scale Missingness**: Only nodes (36 NAs) and differ (46 NAs) need attention.
- **Potential Outliers**: nodes (high node counts), age (younger ages), and time (very short or very long survival).
- **Mostly Categorical/Binary Data**: Many columns are 0/1 or 1/2, which is common in medical datasets (presence/absence, yes/no).
- **Two Records per Patient**: Remember that each patient has two rows (one for recurrence, one for death), which will be important in modeling.


## Step 3: Data Preprocessing and Survival Setup
### 3.1 Convert Data Types
convert binary/categorical variables (sex, obstruct, perfor, adhere, node4, surg, differ, extent) into factors with meaningful labels. This makes the data more readable and suitable for modeling.

```{r}
# Convert sex: 0 = Female, 1 = Male
colon_data$sex <- factor(colon_data$sex, 
                         levels = c(0, 1),
                         labels = c("Female", "Male"))

# Convert obstruct, perfor, adhere: 0 = No, 1 = Yes
colon_data$obstruct <- factor(colon_data$obstruct, 
                              levels = c(0, 1),
                              labels = c("No", "Yes"))
colon_data$perfor <- factor(colon_data$perfor, 
                            levels = c(0, 1),
                            labels = c("No", "Yes"))
colon_data$adhere <- factor(colon_data$adhere, 
                            levels = c(0, 1),
                            labels = c("No", "Yes"))

# Convert node4: 0 = <=4 positive nodes, 1 = >4 positive nodes
colon_data$node4 <- factor(colon_data$node4, 
                           levels = c(0, 1),
                           labels = c("<=4", ">4"))

# Convert surg: 0 = Short, 1 = Long
colon_data$surg <- factor(colon_data$surg, 
                          levels = c(0, 1),
                          labels = c("Short", "Long"))

# Convert differ: 1 = Well, 2 = Moderate, 3 = Poor
colon_data$differ <- factor(colon_data$differ, 
                            levels = c(1, 2, 3),
                            labels = c("Well", "Moderate", "Poor"))

# Convert extent: 1 = Submucosa, 2 = Muscle, 3 = Serosa, 4 = Contiguous structures
colon_data$extent <- factor(colon_data$extent, 
                            levels = c(1, 2, 3, 4),
                            labels = c("Submucosa", "Muscle", "Serosa", "Contiguous"))

```

### 3.2 Handle Missing Data (Multiple Imputation)
use the mice package to impute missing values rather than eliminating rows. This is a more robust approach when dealing with relatively small amounts of missing data.

```{r message=FALSE, warning=FALSE}
### 3.2 Handle Missing Data

# 1. Install/load the 'mice' package if not present
if (!require("mice")) install.packages("mice", dependencies = TRUE)
library(mice)

# 2. Select columns relevant for imputation
#    (Include the variables with missing data + additional predictors)
impute_vars <- c("nodes", "differ", "age", "sex", "obstruct", 
                 "perfor", "adhere", "extent", "node4", "time", "status")

impute_data <- colon_data[, impute_vars]

# 3. Check default methods mice will use
#    For factor variables, 'mice' often uses logistic or polyreg;
#    for numeric, it uses pmm (Predictive Mean Matching) by default.
methods <- make.method(impute_data)
methods

# (Optional) You can manually specify methods, e.g.:
# methods["differ"] <- "polyreg"   # for a factor with >2 levels
# methods["nodes"]  <- "pmm"       # for numeric

# 4. Run multiple imputation
imp <- mice(data = impute_data,
            m = 5,            # number of imputed datasets
            method = methods, # or your customized methods
            maxit = 5,        # number of iterations
            seed = 123)       # for reproducibility

# 5. Choose one of the imputed sets (1 to 5) to create a completed dataset
completed_data <- complete(imp, 1)

# 6. Assign the imputed columns back to your main dataset
colon_data$nodes   <- completed_data$nodes
colon_data$differ  <- completed_data$differ
```
### 3.3 Outlier & Skewness Handling
nodes is heavily skewed. We can log-transform it if we want to treat it as a continuous covariate.

```{r}
### 3.3 Outlier & Skewness Handling

# Create a log-transformed version of 'nodes'
# Adding +1 avoids log(0)
colon_data$log_nodes <- log(colon_data$nodes + 1)


# Function to return a list with lower and upper bounds
get_iqr_bounds <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  return(list(lower = lower_bound, upper = upper_bound))
}

# Let's apply it to 'age', 'time', and 'nodes'
vars_to_check <- c("age","nodes")
bounds_list <- lapply(vars_to_check, function(v) {
  get_iqr_bounds(colon_data[[v]])
})
names(bounds_list) <- vars_to_check

bounds_list
# This will print the lower/upper bound for each variable

### 3.3.2 Flag Outliers

for (v in vars_to_check) {
  lb <- bounds_list[[v]]$lower
  ub <- bounds_list[[v]]$upper
  outlier_col_name <- paste0("outlier_", v)
  
  colon_data[[outlier_col_name]] <- ifelse(colon_data[[v]] < lb | 
                                           colon_data[[v]] > ub, 1, 0)
}

# Check how many outliers in each variable
table(colon_data$outlier_age)
table(colon_data$outlier_time)
table(colon_data$outlier_nodes)

for (v in vars_to_check) {
  lb <- bounds_list[[v]]$lower
  ub <- bounds_list[[v]]$upper
  cap_col_name <- paste0(v, "_wins")
  
  # Winsorize the variable
  colon_data[[cap_col_name]] <- pmin(pmax(colon_data[[v]], lb), ub)
}
# Inspect results
summary(colon_data[, c("nodes","log_nodes","nodes_wins", "age_wins","age")])
```

## Summary of Data Preprocessing

### 1. Multiple Imputation (MICE)
- Created 5 imputed datasets for missing values in `nodes` and `differ`.
- Used one completed dataset (`complete(imp, 1)`) for exploration.
- **Key Point**: For final analysis, pool results across all imputations.

### 2. Outlier Detection (IQR)
- **Age**: Lower bound = 29, Upper bound = 93.
  - 16 outliers (<29); max age = 85 (within bounds).
- **Nodes**: Lower bound = -1, Upper bound = 11.
  - 81 outliers (>11).
- **Interpretation**: Outliers may be clinically valid (e.g., many lymph nodes or young patients).

### 3. Capping & Log Transforms
- **Winsorizing**:
  - `age_wins`: Capped at 29 (min) and 85 (max).
  - `nodes_wins`: Capped at 11 (max).
- **Log Transforms**:
  - `log_nodes`: Range ≈ 0 to 3.53, reducing skewness.
- **Interpretation**: Capping tames extremes; log transforms improve model fit for skewed data.

## Define Censoring Variables

### Decide on Your Survival Endpoint

#### A. Single-Event Analysis (Any Event)
**Goal**: Treat either recurrence or death as the event.
**Method**: Combine both records for each patient so that if a patient has either recurrence or death, we record the earliest event time. Censor if the patient does not experience any event.

#### B. Single-Event Analysis (Death Only)
**Goal**: Model time to death as the sole event.
**Method**: Subset the data to rows where `etype == 2` (death). Censor if the patient has no death record.

#### C. Single-Event Analysis (Recurrence Only)
**Goal**: Model time to recurrence as the sole event.
**Method**: Subset the data to rows where `etype == 1` (recurrence). Censor if the patient has no recurrence record.

#### D. Competing Risks Analysis
**Goal**: Distinguish between recurrence (`etype=1`) and death (`etype=2`) as competing events.
**Method**: Use a competing risks framework (e.g., `cmprsk` or `mstate` package) rather than the standard `Surv()`.

### Approach A: Single-Event Analysis (Any Event)
Goal
- Combine recurrence and death into one “event.”
- If a patient has both recurrence and death, use the earliest occurrence time.
- If a patient has no event, use their latest available time as the censoring time.

```{r}
library(dplyr)
library(survival)

# A) Any Event
colon_any_event <- colon_data %>%
  group_by(id) %>%
  summarize(
    # If the patient has at least one event (status == 1), take the earliest event time
    time = if (any(status == 1)) {
      min(time[status == 1])
    } else {
      # Otherwise, no event => censor at the maximum time
      max(time)
    },
    # Event indicator: 1 if any event, 0 if none
    status = if (any(status == 1)) 1 else 0
  ) %>%
  ungroup()

# Create a Surv object for "any event"
colon_any_event$surv_any <- Surv(colon_any_event$time, colon_any_event$status)

# Example usage:
# fit_any <- survfit(surv_any ~ 1, data = colon_any_event)
# summary(fit_any)


```

### Approach B: Single-Event Analysis (Death Only)
- Goal: Model time to death.
- Recurrence is ignored or treated as if it never “terminates” the subject (i.e., the patient can still die later).
- If no death occurs, censor at the maximum time.

```{r}
# B) Death Only
colon_death <- colon_data %>%
  group_by(id) %>%
  summarize(
    # Earliest time where status=1 and etype=2
    time = if (any(status == 1 & etype == 2)) {
      min(time[status == 1 & etype == 2])
    } else {
      max(time)
    },
    # Event indicator for death
    status = if (any(status == 1 & etype == 2)) 1 else 0
  ) %>%
  ungroup()

colon_death$surv_death <- Surv(colon_death$time, colon_death$status)

# Example usage:
# fit_death <- survfit(surv_death ~ 1, data = colon_death)
# summary(fit_death)

```

### Approach C: Single-Event Analysis (Recurrence Only)
- Goal: Model time to recurrence.
- Death is ignored or treated as a non-informative censoring event (the patient could still theoretically have recurred if they hadn’t died).
- If no recurrence, censor at the maximum time.

```{r}
# C) Recurrence Only
colon_recur <- colon_data %>%
  group_by(id) %>%
  summarize(
    # Earliest time where status=1 and etype=1
    time = if (any(status == 1 & etype == 1)) {
      min(time[status == 1 & etype == 1])
    } else {
      max(time)
    },
    # Event indicator for recurrence
    status = if (any(status == 1 & etype == 1)) 1 else 0
  ) %>%
  ungroup()

colon_recur$surv_recur <- Surv(colon_recur$time, colon_recur$status)

# Example usage:
# fit_recur <- survfit(surv_recur ~ 1, data = colon_recur)
# summary(fit_recur)

```

### Approach D: Competing Risks (Recurrence vs. Death)
- Goal: Treat recurrence (etype=1) and death (etype=2) as distinct, competing events.
- Once one event occurs, the other can’t happen (i.e., no second event).
- Use a specialized package like cmprsk or mstate.

```{r}
# install.packages("cmprsk") # if not installed
library(cmprsk)

```

```{r ,message=FALSE, warning=FALSE}
colon_cr <- colon_data %>%
  group_by(id) %>%
  summarize(
    # If there's any event, pick the earliest time
    time = if (any(status == 1)) {
      min(time[status == 1])
    } else {
      max(time)  # no event => censor at max time
    },
    # Determine which event happened first
    fstatus = case_when(
      # If earliest event was recurrence
      any(status == 1 & etype == 1 & time[status == 1 & etype == 1] == min(time[status == 1])) ~ 1,
      # If earliest event was death
      any(status == 1 & etype == 2 & time[status == 1 & etype == 2] == min(time[status == 1])) ~ 2,
      TRUE ~ 0  # no event
    )
  ) %>%
  ungroup()

# Merge 'rx' or other covariates so we can compare groups
colon_cr <- colon_cr %>%
  left_join(distinct(select(colon_data, id, rx)), by = "id")

# cencode = 0 indicates censored state
# group = colon_cr$rx to compare incidence curves by treatment group
cuminc_result <- cuminc(
  ftime = colon_cr$time,
  fstatus = colon_cr$fstatus,
  group = colon_cr$rx, 
  cencode = 0
)

cuminc_result
# This prints cumulative incidence estimates for each event type (1 or 2),
# stratified by 'rx' if provided.

plot(cuminc_result, xlab = "Days", ylab = "Cumulative Incidence",
     curvlab = c("Recurrence in Obs", "Recurrence in Lev", 
                 "Recurrence in Lev+5FU", 
                 "Death in Obs", "Death in Lev", 
                 "Death in Lev+5FU"))


```
## Understanding the Output

### 1. Gray’s Test
- **stat = 2.04699, p-value = 0.7906895, df = 2**
  - Indicates no statistically significant difference in the cumulative incidence of death among the three groups (Obs, Lev, Lev+5FU).
  - A p-value of ~0.79 is quite high, well above typical significance thresholds (e.g., 0.05).

### 2. Estimates and Variances ($est and $var)
- The table under `$est` shows cumulative incidence values at specific time points (500, 1000, 1500, 2000, 2500, 3000 days).
- Rows (e.g., "Obs 2", "Lev 2", "Lev+5FU 2") correspond to Event = 2 (death) in each treatment group.
- Values in each row are the probability (0 to 1) of having died (Event 2) by that day, accounting for recurrence (Event 1) as a competing risk.
  - For instance, at 3000 days for "Obs 2", you see 0.161825, meaning about 16.2% of patients in the Obs group have died by day 3000 (without having recurred first).

### The Plot: Cumulative Incidence Curves
The plot has six lines in total (Recurrence in Obs/Lev/Lev+5FU and Death in Obs/Lev/Lev+5FU). Focusing on death:
- **Death in Obs** (solid or dashed line labeled “Death in Obs”)
- **Death in Lev**
- **Death in Lev+5FU**

Each line shows how the probability of death (accounting for competing risk of recurrence) evolves over time. By around 3000 days:
- Obs might reach around 16%
- Lev and Lev+5FU might be slightly lower or higher (the table suggests they range around 7–17%)
- Because the p-value is 0.79, these differences are not statistically significant.

### Interpretation & Conclusions

#### No Significant Group Difference for Death
- Gray’s test p-value = 0.79 indicates that none of the three treatments (Obs, Lev, Lev+5FU) show a statistically distinct cumulative incidence of death in this dataset.
- Even if one curve appears a bit higher or lower visually, the difference is too small or too variable to conclude a real difference.

#### Magnitude of Incidence
- By day 3000, the cumulative incidence of death is in the ballpark of 10–16% across groups.
- This suggests that overall, within ~8 years, the probability of dying (without having recurred first) is roughly in that range for each treatment.

#### Recurrence vs. Death
- You also have lines for recurrence in the same plot. Typically, in many colon cancer trials, recurrence rates can be higher than death rates, but that depends on the data.
- If you want to test whether recurrence differs significantly among groups, look for the Event = 1 test or a separate sub-list in your cuminc_result.

#### Clinical Perspective
- If the research question was “Does Lev or Lev+5FU reduce the probability of death compared to Obs?”, these results suggest no clear advantage in terms of cumulative incidence of death.
- However, you should also examine recurrence (Event 1) and/or overall survival to get a complete picture.

## Step 4: Perform Descriptive Analysis
We will perform descriptive analysis by visualizing the distribution of survival times and covariates.

```{r}
# Summary of raw survival times
summary(colon_data$time)
# How many patients had an event vs. censored
table(colon_data$status)        
# Summaries of numeric variables
summary(colon_data$age)
summary(colon_data$nodes)
summary(colon_data$log_nodes)

# Factor (Categorical) Covariates
# Treatment groups
table(colon_data$rx)

# Sex distribution
table(colon_data$sex)

# Tumor differentiation
table(colon_data$differ)

# Extent of disease
table(colon_data$extent)

# Compare a covariate to the event status
table(colon_data$rx, colon_data$status)
table(colon_data$sex, colon_data$status)

# Summary Tables with dplyr or tableone
colon_data %>%
  summarize(
    N = n(),
    MedianTime = median(time, na.rm = TRUE),
    MeanTime = mean(time, na.rm = TRUE),
    EventCount = sum(status == 1),
    CensoredCount = sum(status == 0),
    MeanAge = mean(age, na.rm = TRUE),
    # Add more as needed
  )

library(tableone)

vars <- c("time", "age", "nodes", "sex", "differ")
tab1 <- CreateTableOne(vars = vars, strata = "rx", data = colon_data)
print(tab1, showAllLevels = TRUE)

```
## Overall Dataset Summary

### Total Observations (N): 1858

#### Events vs. Censored:
- **920 events** (status = 1)
- **938 censored** (status = 0)
- This is almost an even split (about 49.5% events).

#### Survival Time (time)
- **Median**: 1855 days (~5.1 years)
- **Mean**: 1537 days (~4.2 years)
- **Range**: 8 to 3329 days
- The mean is lower than the median, indicating a right-skewed distribution.

#### Age
- **Median**: 61 years
- **Range**: 18 to 85 years
- **Mean**: around 59.75 years.

#### Nodes
- **Median**: 2
- **Mean**: 3.66
- **Range**: 0 to 33 (indicating some patients have many positive lymph nodes).

#### Tumor Differentiation (differ)
- **Well**: 191 patients
- **Moderate**: 1360 patients
- **Poor**: 307 patients
- Most tumors are moderately differentiated.

#### Extent of Disease (extent)
- **Submucosa**: 42
- **Muscle**: 212
- **Serosa**: 1518
- **Contiguous**: 86
- The majority (over 80%) are serosa involvement.

## By Treatment Group (rx)
We have three groups: Obs, Lev, and Lev+5FU. A stratified summary shows:

| Group   | N   | Mean Time (SD)     | Mean Age (SD) | Mean Nodes (SD) | % Female | p-values       |
|---------|-----|---------------------|---------------|-----------------|----------|----------------|
| Obs     | 630 | 1440.61 (929.15)    | 59.45 (11.46) | 3.80 (3.71)     | ~52%     | Time < 0.001   |
| Lev     | 620 | 1639.63 (961.71)    | 59.91 (10.89) | 3.69 (3.60)     | ~43%     | Age < 0.001    |
| Lev+5FU | 608 | 1716.50 (102?.??)   | 60.76 (12.25) | 3.62 (3.59)     | ~54%     | Nodes = 0.616  |


### Key Observations

#### Time:
- Lev+5FU has the highest average time (~1716 days). Obs has the lowest (~1441 days).
- p-value < 0.001 indicates a significant difference in survival/follow-up time among the three groups.

#### Age:
- Lev+5FU group is slightly older on average (60.76) compared to Obs (59.45).
- p-value < 0.001 suggests a significant age difference among the groups.

#### Nodes:
- Means range from 3.62 to 3.80; p-value = 0.616 shows no significant difference among groups.

#### Sex:
- ~52% Female in Obs vs. ~43% in Lev vs. ~54% in Lev+5FU. p-value = 0.616 suggests no significant difference among groups.

#### Differentiation (differ):
- p-value = 0.138 indicates no statistically significant difference in tumor differentiation across the three groups.

### Interpretation:
- Time and Age differ significantly among treatment arms, but Nodes, Sex, and Differ do not.
- This might reflect differences in randomization or follow-up among groups (or genuine differences).
- Lev+5FU group has the longest mean follow-up time and slightly older patients on average.

## Event vs. Censoring by Group
From the tables:

- **Obs**: 285 censored, 345 events
- **Lev**: 287 censored, 333 events
- **Lev+5FU**: 366 censored, 242 events

### Interpretation:
- Lev+5FU shows more censored patients than events (366 vs. 242), implying fewer observed events and longer follow-up or better survival.
- Obs has more events (345) than censored (285), possibly indicating shorter survival.
- (Formal testing would require a survival model or log-rank test to confirm if these differences are significant.)

```{r}
##################################################
#  Load Necessary Packages
##################################################
suppressWarnings({
  if (!require("ggplot2")) install.packages("ggplot2", dependencies=TRUE)
  library(ggplot2)
})

##################################################
#  1. Histogram of Survival Time (Any Event)
##################################################
ggplot(colon_data, aes(x = time)) + 
  geom_histogram(binwidth = 100, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Survival Times (Any Event)",
    x = "Days",
    y = "Frequency"
  ) +
  theme_minimal()

##################################################
#  2. Boxplot of Survival Time by Treatment Group
##################################################
ggplot(colon_data, aes(x = rx, y = time, fill = rx)) + 
  geom_boxplot() +
  labs(
    title = "Survival Time by Treatment Group",
    x = "Treatment",
    y = "Days"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  # Hide redundant legend

##################################################
#  3. Boxplot of Survival Time by Tumor Differentiation
##################################################
ggplot(colon_data, aes(x = differ, y = time, fill = differ)) + 
  geom_boxplot() +
  labs(
    title = "Survival Time by Tumor Differentiation",
    x = "Tumor Differentiation",
    y = "Days"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

##################################################
#  4. (Optional) Distribution of Time to Recurrence
##################################################
# Only run this if you have a 'colon_recur' dataset 
# (subset for recurrence event).
ggplot(colon_recur, aes(x = time)) + 
  geom_histogram(binwidth = 100, fill = "blue", color = "black") +
  labs(
    title = "Distribution of Time to Recurrence",
    x = "Days",
    y = "Frequency"
  ) +
  theme_minimal()

##################################################
#  5. (Optional) Distribution of Time to Death
##################################################
# Only run this if you have a 'colon_death' dataset 
# (subset for death event).
ggplot(colon_death, aes(x = time)) + 
  geom_histogram(binwidth = 100, fill = "red", color = "black") +
  labs(
    title = "Distribution of Time to Death",
    x = "Days",
    y = "Frequency"
  ) +
  theme_minimal()

##################################################
#  6. Histogram of Age Distribution
##################################################
ggplot(colon_data, aes(x = age)) + 
  geom_histogram(binwidth = 5, fill = "darkgreen", color = "black") +
  labs(
    title = "Age Distribution",
    x = "Age (Years)",
    y = "Frequency"
  ) +
  theme_minimal()

##################################################
#  7. Distribution of Lymph Nodes
##################################################
ggplot(colon_data, aes(x = nodes)) + 
  geom_histogram(binwidth = 1, fill = "purple", color = "black") +
  labs(
    title = "Distribution of Number of Positive Lymph Nodes",
    x = "Number of Nodes",
    y = "Frequency"
  ) +
  theme_minimal()

ggplot(colon_data, aes(x = log_nodes)) + 
  geom_histogram(binwidth = 0.25, fill = "purple", color = "black") +
  labs(
    title = "Distribution of Log-Transformed Lymph Nodes",
    x = "log(nodes + 1)",
    y = "Frequency"
  ) +
  theme_minimal()


```
## Distribution of Number of Positive Lymph Nodes
- **Shape**: Heavily right-skewed; most patients have 0 to 5 positive nodes, and only a small fraction have >10.
- **Interpretation**: Lymph node involvement is typically low for many patients, but some have high node counts (up to ~30+). This aligns with clinical expectations that while most colon cancer cases have few affected nodes, some present with extensive nodal disease.

## Shape of the Distribution

### After Log-Transforming
- **Fewer Extreme Values**: The high end of the distribution has fewer extreme values.
- **Reduced Skewness**: The data is still somewhat skewed, but less than in the raw count scale.

### Values on the x-axis (log(nodes + 1))
- **log(nodes+1) = 0** → nodes = 0
- **log(nodes+1) = 1** → nodes ≈ 1.72
- **log(nodes+1) = 2** → nodes ≈ 6.39
- **log(nodes+1) = 3** → nodes ≈ 19.09
- **log(nodes+1) = 4** → nodes ≈ 53.60

#### Interpretation
- **Large Spike Around log(nodes+1) = 0–1**: Many patients have 0–2 positive nodes.
- **Few Patients Exceed log(nodes+1) = 3**: Very few patients have more than 19 positive nodes.

## Distribution of Survival Times (Any Event)
- **Overall Range**: Spans from near 0 days up to ~3300 days (over 9 years).
- **Shape**: Appears multi-modal with peaks around 500, 1000, and 2000 days.
- **Interpretation**:
  - A broad spread indicates substantial variability in how quickly patients experience an event (recurrence or death).
  - The multi-modal nature might suggest distinct subgroups of patients or certain clinical follow-up intervals where events cluster.

## Survival Time by Treatment Group (Boxplot)
- **Groups**: Obs (observation), Lev, Lev+5FU.
- **Median Survival**:
  - Obs: Lower median (around ~1000 days).
  - Lev: Intermediate median (~1600 days).
  - Lev+5FU: Higher median (~2000 days).
- **Spread**:
  - Lev+5FU has a wider range (top whisker extends beyond 3000 days).
  - Obs group has many observations below ~1000 days.
- **Interpretation**: Lev+5FU appears to have the longest survival times on average, while Obs has the shortest. However, this is purely descriptive; formal tests (e.g., log-rank or Cox model) would confirm if differences are statistically significant.

## Survival Time by Tumor Differentiation (Boxplot)
- **Categories**: Well, Moderate, Poor.
- **Median Survival**:
  - Well: Highest median (~2000 days).
  - Moderate: Intermediate (~1500 days).
  - Poor: Lowest median (~1000 days).
- **Interpretation**: Patients with well-differentiated tumors show longer survival times on average, whereas poorly differentiated tumors align with shorter times. This aligns with clinical expectations that poorer differentiation often indicates more aggressive disease.

## Distribution of Time to Recurrence
- **Range**: Again up to ~3300 days.
- **Shape**: Multi-modal peaks around 500–600 days and ~2000 days.
- **Interpretation**:
  - Recurrence can occur early (within the first year or two) or later (~5+ years).
  - The presence of a peak around 2000 days might reflect late recurrences, underscoring the need for long-term follow-up.

## Distribution of Time to Death
- **Range**: 0 to ~3300 days.
- **Shape**: Another multi-modal distribution with peaks around 1500–2000 days and smaller clusters elsewhere.
- **Interpretation**:
  - Many deaths occur in the 4–6 year range, but a noticeable number also happen earlier.
  - Indicates colon cancer mortality can be spread across the entire follow-up period, not only in early years.

## Age Distribution
- **Shape**: Roughly bell-shaped (unimodal) with a peak around 60–70 years.
- **Range**: From ~20 to ~85, with a median ~61 (as noted in your summary stats).
- **Interpretation**: Consistent with the typical older demographic for colon cancer, though a small subset of younger patients (<40) also appear.

## Overall Observations & Next Steps
- **Highly Variable Survival Times**: Both for any event and specifically for recurrence or death. Some patients experience events relatively early (~500 days), while others remain event-free for many years.
- **Treatment Differences (Descriptive)**: Lev+5FU group shows higher median survival, while Obs is lower.
- **Tumor Differentiation**: Well-differentiated tumors are associated with longer survival, while poor differentiation correlates with shorter survival.



## Estimate survival functions using Kaplan-Meier curves

Research Question Focus
When conducting survival analysis, it's common to start with the main predictor and time-to-event. Here's a structured approach:

Initial Analysis:

Main Predictor: Focus on the primary variable of interest, such as the treatment group (rx).
Time-to-Event: Analyze the time until the event occurs (e.g., recurrence or death).
Simpler Models:

Kaplan-Meier Curves: Begin with Kaplan-Meier survival curves stratified by the main predictor (e.g., treatment group). This helps to visually assess if there are obvious differences in survival between groups.
Adding Covariates:

Cox Proportional Hazards Model: Once you have a sense of the survival differences, incorporate additional covariates (e.g., age, sex, nodes, obstruct) into a Cox model. This allows for a more comprehensive analysis, adjusting for potential confounders and identifying independent predictors of survival.
```{r}
#  If We already merged data into a single row per patient (Approach A)
# For "any event"
# Surv object:
colon_any_event$surv_any <- Surv(time = colon_any_event$time,
                                  event = colon_any_event$status)
```

```{r}
# Surv object with original data (assuming one row per patient)
colon_data$surv_obj <- Surv(time = colon_data$time, event = colon_data$status)
```

```{r}
# Using the Surv object in colon_data
fit_km_rx <- survfit(surv_obj ~ rx, data = colon_data)

# Print a summary of the fit
summary(fit_km_rx)

```

```{r}
plot(fit_km_rx, 
     xlab = "Days", 
     ylab = "Survival Probability", 
     col = c("blue", "red", "green"), 
     lty = 1:3,
     main = "Kaplan-Meier Curves by Treatment Group")

legend("bottomleft",
       legend = levels(colon_data$rx),
       col = c("blue", "red", "green"),
       lty = 1:3)

```

```{r}
library(survminer)
ggsurvplot(
  fit = fit_km_rx,
  data = colon_data,
  risk.table = TRUE,       # show number at risk table
  pval = TRUE,             # show p-value from log-rank test
  ggtheme = theme_minimal(),
  title = "Kaplan-Meier Curves by Treatment Group",
  xlab = "Days",
  ylab = "Survival Probability"
)

```

```{r}
logrank_test_rx <- survdiff(surv_obj ~ rx, data = colon_data)
logrank_test_rx
```
## Kaplan-Meier Curves by Treatment

### KM Plots (Base R and survminer)
- **Lev+5FU (green/dotted)**: Appears to have the highest survival curve over time.
- **Obs (blue/solid)**: Seems to have the lowest curve.
- **Lev (red/dashed)**: Is in between.

### Survival Probabilities (from the survfit summary)
- For each time point, you see columns for survival, std.err, and 95% CI.
- At early times (like ~50 days), survival is ~0.99 or higher across groups.
- By later times (~3000 days), survival probabilities diverge, with Lev+5FU typically retaining higher survival.

### Interpretation
- Visually, Lev+5FU group experiences fewer (or later) events, while the Obs group experiences more (or earlier) events.
- This pattern suggests Lev+5FU might be more effective at delaying the event (recurrence or death).

## Log-Rank Test
- **Chi-square = 33.6 on 2 degrees of freedom, p = 5e-08 (~0.00000005)**
  - This is a highly significant result (< 0.0001), indicating that at least one survival curve differs from the others.

### From the table:
| Group   | N (Total) | Observed Events | Expected Events |
|---------|-----------|-----------------|-----------------|
| Obs     | 630       | 345             | 299             |
| Lev     | 620       | 333             | 295             |
| Lev+5FU | 608       | 242             | 286             |

- **Obs**: Has more events than expected (345 vs. 299), consistent with lower survival.
- **Lev+5FU**: Has fewer events than expected (242 vs. 286), consistent with better survival.
- **Lev**: Is intermediate.

### Interpretation
- The p-value is extremely small, so we can reject the null hypothesis that all three groups have the same survival function.
- In plain language: Treatment significantly affects time to event (the combined outcome).

## Practical Takeaways
- **Lev+5FU**: Appears to provide a survival advantage, while Obs is least effective.
- **Lev**: Is in between but closer to Obs, based on the shape of the curves and the number of events.

### Manual Pairwise Log-Rank Tests

```{r}
# Pair 1: Obs vs Lev
survdiff(Surv(time, status) ~ rx,
         data = colon_data,
         subset = (rx %in% c("Obs", "Lev")))

# Pair 2: Obs vs Lev+5FU
survdiff(Surv(time, status) ~ rx,
         data = colon_data,
         subset = (rx %in% c("Obs", "Lev+5FU")))

# Pair 3: Lev vs Lev+5FU
survdiff(Surv(time, status) ~ rx,
         data = colon_data,
         subset = (rx %in% c("Lev", "Lev+5FU")))

```
## Pairwise Comparisons of Treatment Groups

### Obs vs Lev
- **p = 0.8**: Suggests no statistical difference in survival curves between Observation alone and Levamisole (Lev).
- **Interpretation**: Visually or clinically, these two groups appear quite similar in how quickly events occur.

### Obs vs Lev+5FU
- **p = 1 × 10⁻⁷**: Indicates a highly significant difference.
- **Interpretation**: Lev+5FU outperforms Obs, implying better survival (fewer/farther events).

### Lev vs Lev+5FU
- **p = 6 × 10⁻⁷**: Also indicates a highly significant difference.
- **Interpretation**: Lev+5FU outperforms Lev, again suggesting better survival.

### Summary
- **Lev+5FU**: Significantly better than both Obs and Lev, suggesting it provides a survival advantage.
- **Obs and Lev**: Not significantly different from each other, indicating similar survival outcomes.

These results highlight the effectiveness of Lev+5FU in improving survival compared to the other treatment groups.

###########################################################################################
## Kaplan-Meier Curves (Time to Death)

```{r}
# 1) Summarize time/status for death
colon_death <- colon_data %>%
  group_by(id) %>%
  summarize(
    time = if (any(status == 1 & etype == 2)) {
      min(time[status == 1 & etype == 2])
    } else {
      max(time)
    },
    status = if (any(status == 1 & etype == 2)) 1 else 0
  ) %>%
  ungroup()

# 2) Merge 'rx' from 'colon_data'
colon_death <- colon_death %>%
  left_join(
    select(colon_data, id, rx),
    by = "id"
  ) %>%
  distinct(id, .keep_all = TRUE)

# Now 'colon_death' should have columns: id, time, status, rx, ...
str(colon_death)

```


```{r}
# Surv object for time to death
colon_death$surv_death <- Surv(time = colon_death$time,
                               event = colon_death$status)

```

```{r}
# Suppose you want to compare survival by rx (Obs, Lev, Lev+5FU)
fit_km_death <- survfit(surv_death ~ rx, data = colon_death)

# Print a summary
summary(fit_km_death)

```


```{r}
plot(fit_km_death,
     xlab = "Days",
     ylab = "Survival Probability (Time to Death)",
     col = c("blue", "red", "green"),
     lty = 1:3,
     main = "Kaplan-Meier Curves: Time to Death by Treatment")

legend("bottomleft",
       legend = levels(colon_death$rx),
       col = c("blue", "red", "green"),
       lty = 1:3)


ggsurvplot(
  fit = fit_km_death,
  data = colon_death,
  risk.table = TRUE,
  pval = TRUE,
  xlab = "Days",
  ylab = "Survival Probability (Death)",
  title = "Kaplan-Meier: Time to Death by Treatment",
  ggtheme = theme_minimal()
)

```

```{r}
logrank_death <- survdiff(surv_death ~ rx, data = colon_death)
logrank_death

```
```{r}
survdiff(Surv(time, status) ~ rx, data = colon_death,
         subset = (rx %in% c("Obs", "Lev")))
survdiff(Surv(time, status) ~ rx, data = colon_death,
         subset = (rx %in% c("Obs", "Lev+5FU")))
survdiff(Surv(time, status) ~ rx, data = colon_death,
         subset = (rx %in% c("Lev", "Lev+5FU")))
```
## Kaplan-Meier Curves (Time to Death)

### 1. Visual Ranking of Survival Curves
- **Lev+5FU (green/dotted)**: Remains the highest curve (best survival), meaning fewer patients die (or they die later).
- **Obs (blue/solid)** and **Lev (red/dashed)**: Are closer together. Sometimes one is slightly above the other, but they appear more similar than in the any-event analysis.

### Survival Probabilities Over Time
- **Early on (first 500–1000 days)**: Survival probabilities are quite high for all three groups.
- **By ~3000 days**: The curves have separated more clearly, with Lev+5FU maintaining the highest survival and Obs vs. Lev looking fairly close.

### 2. Log-Rank Test for Time to Death
- **Chi-square = 11.7, p = 0.003**: Indicates a statistically significant difference in time to death among the three groups.

#### Interpretation of Observed vs. Expected:
- **Obs**: 168 deaths observed vs. 148 expected, indicating more deaths than expected under a null (no-difference) scenario.
- **Lev**: Also has more observed than expected.
- **Lev+5FU**: Presumably has fewer observed than expected, aligning with better survival.

### 3. Comparison with “Any Event” Results
- **Any Event Analysis**: Had an extremely low p-value (~10⁻⁷) because Lev+5FU was drastically better for recurrence and death combined.
- **Time to Death**: Yields p = 0.003—still significant but less extreme.
- **Implication**: The difference among treatments is still real for death alone, but the magnitude of separation is somewhat smaller than in the any-event scenario. This suggests Lev+5FU provides a strong advantage particularly for recurrence prevention, though it also improves overall survival compared to the others.

### 4. Do Obs and Lev Differ?
- The log-rank test is global, so we only know that at least one pair differs.
- **Visually**: Obs and Lev appear close. You can do pairwise tests (subsetting to (Obs, Lev) only) to see if that difference is statistically significant.
- **If the p-value is large**: It means Obs vs. Lev are not significantly different in terms of time to death.

#################################################################################
##  Kaplan-Meier Curves (Time to Recurrence)

```{r}
# Step 1: Summarize time/status for recurrence
colon_recur <- colon_data %>%
  group_by(id) %>%
  summarize(
    # Earliest time if (status=1 & etype=1), otherwise max time
    time = if (any(status == 1 & etype == 1)) {
      min(time[status == 1 & etype == 1])
    } else {
      max(time)
    },
    # Recurrence indicator: 1 if event occurred, else 0
    status = if (any(status == 1 & etype == 1)) 1 else 0
  ) %>%
  ungroup()

# Step 2: Merge `rx` from the original data
colon_recur <- colon_recur %>%
  left_join(
    select(colon_data, id, rx),
    by = "id"
  ) %>%
  distinct(id, .keep_all = TRUE)

# Check structure
str(colon_recur)
head(colon_recur)

```

```{r}
# Create a Surv object for time to recurrence
colon_recur$surv_recur <- Surv(time = colon_recur$time,
                               event = colon_recur$status)

```

```{r}
fit_km_recur <- survfit(surv_recur ~ rx, data = colon_recur)

# View a textual summary of survival estimates
summary(fit_km_recur)

```

```{r}
plot(fit_km_recur,
     xlab = "Days",
     ylab = "Survival Probability (Recurrence)",
     col = c("blue", "red", "green"),
     lty = 1:3,
     main = "Kaplan-Meier: Time to Recurrence by Treatment")

legend("bottomleft",
       legend = levels(colon_recur$rx),
       col = c("blue", "red", "green"),
       lty = 1:3)

ggsurvplot(
  fit = fit_km_recur,
  data = colon_recur,
  risk.table = TRUE,
  pval = TRUE,  # displays log-rank p-value
  xlab = "Days",
  ylab = "Survival Probability (Recurrence)",
  title = "Kaplan-Meier: Time to Recurrence by Treatment",
  ggtheme = theme_minimal()
)

```

```{r}
logrank_recur <- survdiff(surv_recur ~ rx, data = colon_recur)
logrank_recur
```
```{r}
# Obs vs Lev
survdiff(Surv(time, status) ~ rx, data = colon_recur,
         subset = (rx %in% c("Obs", "Lev")))
# Obs vs Lev+5FU
survdiff(Surv(time, status) ~ rx, data = colon_recur,
         subset = (rx %in% c("Obs", "Lev+5FU")))
# Lev vs Lev+5FU
survdiff(Surv(time, status) ~ rx, data = colon_recur,
         subset = (rx %in% c("Lev", "Lev+5FU")))

```
## Kaplan-Meier Curves (Time to Recurrence)

### 1. Visual Ranking
- **Lev+5FU (green/dotted)**: Shows the highest curve, meaning fewer patients recurred (or they recurred later).
- **Obs (blue/solid)** and **Lev (red/dashed)**: Appear lower, indicating earlier or more frequent recurrences.

### Recurrence-Free Survival Probability
- **Early on (first 500–1000 days)**: Survival probabilities (i.e., no recurrence) are fairly high in all groups.
- **By ~3000 days**: Lev+5FU remains clearly above 0.40, while Obs and Lev dip lower, indicating a higher proportion of patients in the Lev+5FU group remain recurrence-free over time.

### Interpretation
- **Lev+5FU**: Appears to be most effective at delaying or preventing recurrence.
- **Obs vs. Lev**: Are closer to each other, though you’d need pairwise tests to see if that difference is significant.

## Log-Rank Test for Time to Recurrence
- **Chi-Square = 23.1, p = 1e-05 (~0.00001)**
  - This is highly significant, indicating that at least one group’s recurrence-free survival differs from the others.

### Observed vs. Expected
- **Obs**: 177 recurrences observed vs. 151 expected → more recurrences than expected (worse outcome).
- **Lev**: 172 vs. 149 → also more recurrences than expected (worse).
- **Lev+5FU**: 119 vs. 169 → many fewer recurrences than expected (better).

### Interpretation
- The large difference is primarily driven by Lev+5FU performing significantly better (far fewer recurrences than expected).
- Obs and Lev have higher observed recurrences than expected, consistent with worse recurrence-free survival.

## Clinical/Practical Meaning
- **Lev+5FU**: Strong evidence it reduces recurrence compared to the other two treatments.
- **Obs vs. Lev**: The global log-rank test can’t tell you if they differ significantly from each other; you’d do pairwise tests (e.g., subsetting `rx %in% c("Obs", "Lev")`) to confirm.

## Pairwise Log-Rank Results
we ran three separate `survdiff` commands with subset to isolate each pair of groups:

### Obs vs Lev
- **χ² = 0.1, p = 0.9**
  - No significant difference in recurrence-free survival between Obs and Lev.

### Obs vs Lev+5FU
- **χ² = 19.1, p = 1 × 10⁻⁵**
  - Highly significant difference: Lev+5FU is better (fewer recurrences).

### Lev vs Lev+5FU
- **χ² = 17.7, p = 3 × 10⁻⁵**
  - Highly significant difference: Lev+5FU is better than Lev.

## Interpretation
- **Obs vs Lev**:
  - **p = 0.9**: They do not differ significantly in terms of time to recurrence.
  - **Practical terms**: Levamisole alone does not provide a statistically significant benefit over observation for preventing recurrence.

- **Obs vs Lev+5FU**:
  - **p = 1e-05**: Strong evidence that Lev+5FU significantly reduces recurrence compared to observation alone.

- **Lev vs Lev+5FU**:
  - **p = 3e-05**: Also highly significant; Lev+5FU is much better than Lev alone at preventing recurrence.

### Overall Pattern
- **Lev+5FU**: Stands out as the only regimen that significantly delays or reduces recurrence.
- **Obs and Lev**: Are indistinguishable in this dataset for recurrence outcomes (i.e., no significant difference).

## Clinical/Practical Meaning
- **Lev+5FU**: Is clearly the optimal regimen for preventing recurrence.
- **Lev alone**: Does not appear to outperform Obs for recurrence-free survival.
- These findings confirm the global test result (p < 0.0001) was driven by Lev+5FU being significantly superior to both other arms.


###################################################################################################

## Cox Proportional Hazards Modeling

## Time to Any Event

### What It Does
- Combines recurrence and death into a single outcome.
- If a patient has either event, they’re considered to have “failed” at that time.

### Pros
- Simple summary of overall treatment failure (recurrence or death).
- Captures the earliest time a patient experiences any negative outcome.

### Cons
- No distinction between recurrence and death; you lose details about which event happened.
- If you care about overall survival specifically, or recurrence specifically, combining them can blur those details.

### When to Choose It
- We want a broad measure of “treatment failure,” and either recurrence or death is equally critical as an endpoint.
- Ideal if your research question is: “Does this treatment reduce the chance of any negative event (recurrence or death)?”

## Time to Death (Overall Survival)

### What It Does
- Only death counts as an event.
- Recurrence is ignored or treated as censored (the patient is still alive and at risk of eventually dying).

### Pros
- Clinically straightforward if your main endpoint is overall survival.
- Often the gold standard in many oncology trials: does the treatment prolong life?

### Cons
- Doesn’t capture recurrence patterns, which can be clinically important.
- A patient could recur early but still live for many years, and you’d miss that early recurrence in a time-to-death analysis.

### When to Choose It
- our primary question is about mortality: “Does the treatment reduce the risk of death?”
- Overall survival is often used in regulatory approvals and major clinical endpoints.

## Time to Recurrence

### What It Does
- Recurrence is the event; death is treated as censored (the patient is no longer at risk of recurrence if they die first).

### Pros
- Lets you focus on disease progression.
- Particularly important if your key question is about recurrence-free survival or relapse rates.

### Cons
- If patients die before recurrence, they’re censored, which can introduce bias if death is related to the same disease process.
- Does not capture overall survival. A treatment might reduce recurrence but not necessarily extend life (or vice versa).

### When to Choose It
- We care about tumor relapse specifically: “Does the treatment delay or prevent the return of cancer?”
- Recurrence is a clinically important endpoint (e.g., new metastatic lesions).


we choose the first approach !



## Select Covariates (with Reasons)

### 2.1 Commonly Included Predictors in Colon Cancer

#### rx (Treatment)
- **Reason**: Primary variable of interest. You want to see how each treatment arm (Obs, Lev, Lev+5FU) affects survival.

#### age (Numeric)
- **Reason**: Age is a universal risk factor; older patients often have worse outcomes.

#### sex (Male/Female)
- **Reason**: Sometimes survival differs by sex in colon cancer, though not always strongly.

#### nodes (Number of positive lymph nodes)
- **Reason**: Very strong prognostic factor in colon cancer. Often right-skewed, so consider log transform if needed.

#### differ (Tumor differentiation: Well, Moderate, Poor)
- **Reason**: Histological grade can strongly impact prognosis.

#### extent (Extent of local spread: Submucosa, Muscle, Serosa, Contiguous)
- **Reason**: Deeper invasion often means higher risk.

### 2.2 Optional or Secondary Predictors

#### obstruct, perfor, adhere (0/1)
- **Reason**: May indicate more complicated surgeries or advanced local disease. Sometimes less commonly included, but can be tested.

#### surg (Short vs. Long interval from surgery to registration)
- **Reason**: Might matter for follow-up patterns.

#### node4 (>4 positive nodes)
- **Reason**: A simpler, binary version of nodes. But if you already have the numeric nodes, node4 can be redundant.

### covariate list: 

```{r}
# Example covariates
vars <- c("rx", "sex", "age", "log_nodes", "differ", "extent", 
          "obstruct", "perfor", "adhere", "surg")
```

###  Create the Surv Object

```{r}

##################################################
# 1. Load Packages
##################################################
# install.packages(c("dplyr", "survival"))
library(dplyr)
library(survival)

##################################################
# 2. Summarize to Create "colon_any_event"
#    One Row per Patient, Earliest Event Time
##################################################
colon_any_event <- colon_data %>%
  group_by(id) %>%
  summarize(
    # If a patient has any event (status == 1),
    # take the earliest time of that event.
    # Otherwise, if no event, use max time (censored).
    time = if (any(status == 1)) {
      min(time[status == 1])
    } else {
      max(time)
    },
    status = if (any(status == 1)) 1 else 0
  ) %>%
  ungroup()

##################################################
# 3. Left Join Covariates from Original Data
##################################################
# Select the covariates you want to keep:
# e.g., rx, age, sex, nodes, differ, extent, obstruct, perfor, adhere, surg
colon_any_event <- colon_any_event %>%
  left_join(
    select(colon_data, 
           id, rx, age, sex, log_nodes, nodes, differ, extent,
           obstruct, perfor, adhere, surg),
    by = "id"
  ) %>%
  # If 'colon_data' had multiple rows per id,
  # distinct() ensures we keep only one row per patient
  distinct(id, .keep_all = TRUE)

# Check structure to confirm covariates are present
str(colon_any_event)
head(colon_any_event)

##################################################
# 4. Create the Surv Object (Time to Any Event)
##################################################
colon_any_event$surv_any <- Surv(colon_any_event$time,
                                 colon_any_event$status)

```



### Univariable Cox Models in a Loop

```{r}
# Function to run univ Cox and extract tidy results
univ_results <- lapply(vars, function(v) {
  # Build a formula like: Surv(time, status) ~ var
  form <- as.formula(paste("surv_any ~", v))
  
  # Fit univariable Cox model
  fit <- coxph(form, data = colon_any_event)
  
  # Tidy results with broom
  res <- tidy(fit, conf.int = TRUE) %>%
    mutate(variable = v)  # keep track of which var
  return(res)
})

# Combine into one data frame
univ_results_df <- do.call(rbind, univ_results)

# Add columns for Hazard Ratio (HR) and 95% CI
univ_results_df <- univ_results_df %>%
  mutate(HR = exp(estimate),
         HR_lower = exp(conf.low),
         HR_upper = exp(conf.high))
```
### Visualize Univariable Results with ggplot2

```{r}
# Print table of univ results
print(univ_results_df)

# Forest plot with ggplot2
ggplot(univ_results_df, aes(x = variable, 
                            y = HR, 
                            ymin = HR_lower, 
                            ymax = HR_upper)) +
  geom_pointrange() +
  geom_hline(yintercept = 1, linetype = 2, color = "red") +
  coord_flip() +
  scale_y_log10() +
  labs(
    x = "Covariate",
    y = "Hazard Ratio (log scale)",
    title = "Univariable Cox PH - Time to Any Event"
  ) +
  theme_minimal()
```
## Univariate Analysis Findings

### From the Forest Plot and Univariable Table

#### Covariates with p < 0.05:
- **log_nodes**: Highly significant (HR > 1).
- **rxLev+5FU**: Significantly reduces hazard (HR < 1).
- **Possibly obstruct borderline or surg borderline**: Depending on your exact p-values.

#### Covariates with p ≥ 0.05:
- Some, like **sex**, **perfor**, or **adhere**, may not be statistically significant in univariable testing.
- Others, like **extent** or **differ**, might have partial significance depending on how the factor levels are compared (some levels can be borderline).

### Interpretation
- **log_nodes**: Stands out as a strong predictor; more nodes → higher hazard.
- **rxLev+5FU**: Consistently shows hazard < 1, meaning improved survival relative to rxObs.

### Code to Filter Significant Variables 

```{r}
# Suppose your univ results are in univ_results_df
sig_vars <- univ_results_df %>%
  filter(p.value < 0.05) %>%
  pull(variable)

sig_vars

```

### Build the Initial Multivariable Cox Model

```{r}
# including "rx"       "log_nodes"    "differ"   "extent"   "extent"   "obstruct" "adhere"   "surg"  
cox_model <- coxph(
  Surv(time, status) ~ rx + obstruct + log_nodes + differ + extent + adhere + surg,
  data = colon_any_event
)
summary(cox_model)

```
## Building the Multivariable Cox Model

### Combined Covariates:
- **rx** (main treatment factor)
- **log_nodes** (since it’s strongly significant)
- **obstruct**, **differ**, **extent**, **adhere**, **surg** (some are borderline or clinically relevant)

### Significant Variables (p < 0.05):
- **rxLev+5FU**: HR ≈ 0.62 → ~38% reduction in hazard vs. rxObs.
- **log_nodes**: HR ≈ 1.91 → Each log-unit increase in nodes nearly doubles the hazard.
- **extentContiguous**: HR ≈ 2.88 → Markedly higher hazard for contiguous spread vs. baseline category (likely Submucosa or whichever is reference).
- **surgLong**: HR ≈ 1.31 → “Long” interval from surgery to registration is associated with ~31% higher hazard.

### Borderline:
- **obstructYes**: p ~ 0.055.
- **differPoor**: p ~ 0.10.

### Not Significant:
- **rxLev**: p = 0.63 → Not significantly different from rxObs.
- **differModerate**, **extentMuscle**, **extentSerosa**, **adhereYes**.

### Interpretation:
- **rxLev+5FU**: Shows a significant reduction in hazard, indicating improved survival compared to rxObs.
- **log_nodes**: Strong predictor; more nodes → higher hazard.
- **extentContiguous**: Indicates a markedly higher hazard for contiguous spread.
- **surgLong**: Suggests a higher hazard associated with a longer interval from surgery to registration.
- **Borderline and Not Significant Variables**: These may not have a strong impact on survival in this model, but could be clinically relevant in other contexts

## Check Cox PH Assumption

```{r}
ph_test <- cox.zph(cox_model)
ph_test
plot(ph_test)
```
## Proportional Hazards Assumption Violation

### Global Test
- **p = 5.0e-06**: Overall, the model violates proportional hazards.

### Covariates with p < 0.05 in PH Test:
- **obstruct** (p = 0.0070)
- **log_nodes** (p = 0.0073)
- **differ** (p = 1.1e-06)

This means obstruct, log_nodes, and differ show evidence that their hazard ratios change over time. The plotted scaled Schoenfeld residuals (like for log_nodes or differ) might show a non-horizontal trend.

### Interpretation
- The PH assumption is violated for these covariates.

### Potential Solutions:
- **Stratify on obstruct or differ**: If they’re key categorical variables.
- **Time-varying approach for log_nodes**: If it strongly changes over time.
- **Check if recoding or grouping factor levels helps**: Simplifying categories might reduce violations.
- **Evaluate whether you still want to keep them in the model**: If the violation is moderate, consider the trade-offs.

## Visual Inspection of Residual Plots

### Schoenfeld Residual Plots for Each Variable

#### rx
- **Observation**: The fitted line is fairly flat around 0.
- **Interpretation**: No major PH violation for rx.

#### obstruct
- **Observation**: Some drift over time.
- **Interpretation**: This aligns with the p = 0.007 in the PH test.

#### log_nodes
- **Observation**: Slight slope or curvature in the fitted line.
- **Interpretation**: This might be mild or moderate.

#### differ
- **Observation**: Possibly a noticeable downward trend.
- **Interpretation**: Indicates time-dependent effects.

#### extent
- **Observation**: Fairly horizontal.
- **Interpretation**: No big violation.

#### adhere, surg
- **Observation**: Also appear fairly stable.
- **Interpretation**: No major PH violations.

### Summary
- **rx**: No major PH violation.
- **obstruct**: Shows some drift, indicating a potential violation.
- **log_nodes**: Mild to moderate violation.
- **differ**: Indicates time-dependent effects.
- **extent, adhere, surg**: No significant violations.

##  Check Linearity for Numeric Variables

```{r}
resid_mart <- residuals(cox_model, type = "martingale")

# For 'nodes'
plot(colon_any_event$nodes, resid_mart,
     xlab = "nodes", ylab = "Martingale Residuals")
abline(h = 0, col = "red")

# For 'age'
plot(colon_any_event$age, resid_mart,
     xlab = "age", ylab = "Martingale Residuals")
abline(h = 0, col = "red")

```
## Nodes Plot

### Observations:
- Many points are near 0 for node counts under ~10, then some outliers for high node counts (20–30+).
- Possibly a mild negative slope for low node values, but the main cluster is near 0.
- High node counts (above ~10) are relatively sparse but some negative residuals.

### Interpretation:
- The overall pattern doesn’t show a strong curve, but there is a cluster effect from so many low node counts.
- we’ve already log-transformed nodes (log_nodes), which typically helps with the skew. This plot suggests it’s mostly okay, though you have a few outliers at high node counts.
- No glaring “U-shape” or strong departure from linearity is evident. The log transform is likely addressing the skew effectively.

## Age Plot

### Observations:
- Points are spread from age ~20 to 85, with a cluster around 50–70.
- The red line (loess or a simple horizontal) is fairly flat, maybe a slight downward slope for older ages.
- No strong curvature or obvious pattern.

### Interpretation:
- No major violation of linearity for age. A gentle slope might exist, but it’s not pronounced.
- If clinically you suspect age might have a threshold effect (e.g., <50 vs. >70), you could explore a piecewise approach. But from this plot, age seems reasonably linear.

####################################################################################

## Evaluate Model Performance (Harrell’s C-Index)

```{r}
# Suppose cox_model is  final model
summary(cox_model)$concordance
```

##  Alternative: survcomp or pec Packages

```{r}
library(survcomp)
# x is your predicted risk from the Cox model (e.g., linear predictor)
x <- predict(cox_model, type = "lp")
# Surv object
S <- with(colon_any_event, Surv(time, status))

cindex_result <- concordance.index(x, surv.time = S[,1], surv.event = S[,2],
                                   method = "noether")
cindex_result$c.index
cindex_result$lower
cindex_result$upper

```
## Perform Cross-Validation or Bootstrapping
Bootstrapping for Robustness
One common approach is to bootstrap the dataset multiple times and refit the Cox model to estimate the variability of coefficients and the C-index. Here’s a simplified example using the boot package:

```{r, warning=FALSE}
# install.packages("boot")
library(boot)
library(survival)

# 1) Define a function that fits the Cox model and returns the C-index
cox_boot <- function(data, indices) {
  d <- data[indices, ]  # bootstrap sample
  fit <- coxph(Surv(time, status) ~ rx + log_nodes + differ + extent + adhere + surg,
               data = d)
  # Return the C-index from summary
  return(summary(fit)$concordance[1]) 
}

# 2) Run boot
set.seed(123)
boot_res <- boot(data = colon_any_event, statistic = cox_boot, R = 1000) 
# R=1000 bootstrap replications

# 3) Inspect results
boot_res
# 't' contains the bootstrap estimates of C-index
# you can compute mean, sd, etc.

mean(boot_res$t)   # average C-index across bootstraps
sd(boot_res$t)     # standard deviation

```
##  Cross-Validation Approach
we could do k-fold cross-validation (e.g., caret or cvTools) to split data into folds, train on k-1 folds, and test on the remaining fold. Then compute the C-index or partial log-likelihood on the test fold. Summarize across folds for an estimate of out-of-sample performance.
```{r}
cox_cv_cindex <- function(data, formula, k = 5, seed = 123) {
  set.seed(seed)
  
  n <- nrow(data)
  # Randomly assign each row to a fold (1..k)
  folds <- sample(rep(1:k, length.out = n))
  
  cindex_values <- numeric(k)  # store c-index for each fold
  
  for (i in 1:k) {
    # Split into training vs. test
    train_data <- data[folds != i, ]
    test_data  <- data[folds == i, ]
    
    # Fit Cox model on training
    fit <- coxph(formula, data = train_data)
    
    # Predict linear predictor (risk score) on test
    lp_test <- predict(fit, newdata = test_data, type = "lp")
    
    # Compute C-index for test set
    # Survcomp expects vectors: times, events, and predicted scores
    cindex_result <- concordance.index(
      x          = lp_test,                    # predicted risk
      surv.time  = test_data$time,             # actual survival time
      surv.event = test_data$status,           # actual event
      method     = "noether"                   # or "uno", "harrell", etc.
    )
    
    cindex_values[i] <- cindex_result$c.index
  }
  
  # Return average C-index and all fold values
  list(
    mean_cindex = mean(cindex_values),
    sd_cindex   = sd(cindex_values),
    cindex_each_fold = cindex_values
  )
}

```

```{r}
# Define your final formula
cox_formula <- Surv(time, status) ~ rx + log_nodes + differ + extent + adhere + surg

# Run 5-fold CV
cv_results <- cox_cv_cindex(data = colon_any_event, formula = cox_formula, k = 5, seed = 123)

cv_results$mean_cindex
cv_results$sd_cindex
cv_results$cindex_each_fold
```
## 1. In-Sample C-Index from summary(cox_model)
**Value:** ~0.6724 (SE ~0.0117).

**Interpretation:** On your full dataset, the model can correctly rank pairs of patients ~67% of the time. This is a moderate level of discrimination, typical in many clinical models.

## 2. C-Index via survcomp
**Value:** ~0.6734 (95% CI: 0.6504–0.6964).

**Interpretation:** This matches closely with the in-sample summary. The confidence interval indicates you can be fairly confident the true c-index lies in the ~0.65–0.70 range.

## 3. Bootstrapped C-Index
**Value:** ~0.6699 (bias ~0.0027, std. error ~0.012).

**Interpretation:** When you resample (with replacement) 1000 times, the average c-index is ~0.67, reinforcing that your model’s discrimination is stable and not the product of random chance. The std. error is ~0.012, consistent with the original SE ~0.0117.

## 4. 5-Fold Cross-Validation C-Index
**Mean:** ~0.6582  
**SD:** ~0.0137  
**Fold-by-Fold:** [0.6885, 0.6514, 0.6303, 0.7091, 0.6357]

**Interpretation:**
- **Slightly Lower Than In-Sample (0.6724 → 0.6582):** A small drop is expected due to out-of-sample testing. This indicates mild overfitting, but the difference is not large.
- **Fold Variability:** The c-index ranges from ~0.63 to ~0.71 across folds, giving a standard deviation ~0.014. This is moderate and suggests the model’s performance is fairly consistent across data splits.
- **Conclusion:** A cross-validated c-index around 0.66 (±0.01) confirms moderate discrimination in an out-of-sample setting.

## 5. Overall Conclusions
- **Consistent Performance:** All methods (in-sample, survcomp, bootstrapping, cross-validation) converge around 0.66–0.67 for the c-index.
- **Moderate Discrimination:** A c-index in the mid-0.60s is common in medical prognostic models, meaning the model is useful but not perfect.
- **Model Stability:**
  - Bootstrapping shows minimal bias and an SE ~0.012, so the estimate is stable.
  - Cross-Validation c-index is only slightly lower than in-sample, indicating limited overfitting.
- **Practical Interpretation:** Your Time to Any Event Cox model ranks patients’ risk in a moderate way. In a clinical context, this can still provide valuable prognostic insight, but the model is not extremely high in discrimination.
################################################################################

## Interpret Model Coefficients and Survival Curves

```{r}
summary(cox_model)
```
## Covariate-by-Covariate Interpretation

### 1. rxLev
**HR = 0.9505, p = 0.628**  
Not significantly different from 1. This suggests Lev alone is not significantly better or worse than the reference treatment (likely rx=Obs) for preventing the event.

### 2. rxLev+5FU
**HR = 0.6181, p = 2.29e-05 (highly significant)**  
Interpreted as ~38% reduction in hazard relative to the reference group (Obs).  
In plain language: “Patients on Lev+5FU have significantly lower risk (by ~38%) than those on the reference treatment.”

### 3. obstructYes
**HR = 1.2337, p = 0.055 (borderline significance)**  
Suggests a 23% increase in hazard for patients with obstruction vs. those without, but only borderline significant (p ~ 0.055).  
In plain language: “Having an obstructed colon might slightly raise the risk of recurrence/death, but the evidence is borderline.”

### 4. log_nodes
**HR = 1.9140, p < 2e-16 (very highly significant)**  
A 1-unit increase in log(nodes+1) nearly doubles the hazard.  
In simpler terms: “If (nodes+1) goes up by a factor of $$e^1 = 2.718$$, hazard almost doubles.”  
In plain language: “The number of positive lymph nodes is a strong predictor—more nodes means a substantially higher risk.”

### 5. differModerate
**HR = 1.0124, p = 0.938**  
No significant difference from the reference (likely differ=Well).  
“Moderate differentiation” doesn’t show a clear effect vs. well-differentiated tumors.

### 6. differPoor
**HR = 1.3543, p = 0.099 (borderline significance)**  
Suggests a ~35% increase in hazard vs. well-differentiated, but p ~ 0.10 is borderline.  
In plain language: “Poorly differentiated tumors may raise risk, but it’s not firmly significant.”

### 7. extentMuscle
**HR = 1.0550, p = 0.903**  
Not significant. Minimal difference from reference category (likely Submucosa).

### 8. extentSerosa
**HR = 1.6966, p = 0.202**  
Suggests ~70% higher hazard, but p=0.20 is not significant. No strong evidence that “Serosa” is worse than the reference category.

### 9. extentContiguous
**HR = 2.8847, p = 0.0188 (significant)**  
~2.9x hazard vs. the reference.  
In plain language: “Tumor extending to contiguous structures significantly increases the risk (almost triple).”

### 10. adhereYes
**HR = 1.2024, p = 0.129**  
Not significant (p=0.13). Possibly ~20% higher hazard, but not statistically confirmed.

### 11. surgLong
**HR = 1.3080, p = 0.0057 (significant)**  
~31% higher hazard if surgery-to-registration interval is “Long” vs. “Short.”  
In plain language: “Patients who had a longer gap between surgery and registration face a higher risk.”

## Overall Model Summary
- **Concordance (C-index) ~ 0.672:** The model’s discrimination is moderate (can rank patients’ risk ~67% of the time correctly).
- **Likelihood Ratio / Wald / Score tests:** All highly significant (p < 2e-16), meaning the model overall is very unlikely to be random


## Visualizing Survival Curves from a Cox Model
```{r}
newdata_rx <- data.frame(
  rx = c("Obs", "Lev", "Lev+5FU"),
  obstruct = "No",
  log_nodes = mean(colon_any_event$log_nodes, na.rm=TRUE),
  differ = "Moderate",
  extent = "Serosa",
  adhere = "No",
  surg = "Short"
)

fit_surv <- survfit(cox_model, newdata = newdata_rx)

plot(fit_surv, col=c("blue","red","green"), lty=1:3,
     xlab="Days", ylab="Survival Probability",
     main="Adjusted Survival by Treatment")
legend("bottomleft", legend=c("Obs","Lev","Lev+5FU"),
       col=c("blue","red","green"), lty=1:3)

```


## Conclusion

### Data Preparation & Exploration
- We began by importing the Colon dataset, checking for missing values (notably in nodes and differ) and addressing them via imputation.
- Outlier checks (e.g., nodes, age) led to transformations (log_nodes) to handle skew.
- Basic Kaplan-Meier analyses indicated Lev+5FU was beneficial compared to Obs, while Lev alone was not clearly different.

### Cox PH Modeling
- Univariable tests identified rx=Lev+5FU, log_nodes, and certain tumor extent levels as strong predictors.
- The final Cox model included:
  Surv(time, status) ∼ rx + obstruct + log_nodes + differ + extent + adhere + surg.
- Key significant effects:
  - **rx=Lev+5FU:** Hazard ratio ~0.62 (p<0.0001), a ~38% hazard reduction vs. observation.
  - **log_nodes:** HR ~1.91, the strongest risk factor—more positive nodes drastically increase hazard.
  - **extentContiguous:** ~2.9x hazard vs. baseline.
  - **surgLong:** ~31% higher hazard vs. surgShort.

### Model Assumption Checks & Validation
- **PH Tests:** Showed potential violations for obstruct, log_nodes, differ, suggesting mild time-varying effects or need for stratification.
- **Performance:**
  - In-sample C-index: ~0.67, indicating moderate discrimination.
  - Bootstrapping: Confirmed stability with minimal bias and SE ~0.012.
  - 5-Fold Cross-Validation: Mean C-index ~0.66, close to in-sample, implying limited overfitting.

### Adjusted Survival Curves
- Visualizing survival by rx with other covariates at average levels showed a notably higher curve for Lev+5FU, confirming its protective effect.
- Lev alone tracked close to observation, matching the non-significant difference in hazard.

### Overall Findings
- **Lev+5FU** stands out as the most effective regimen in this dataset, significantly improving survival compared to observation or Lev alone.
- **Number of nodes (log_nodes)** remains the dominant risk factor, highlighting the critical prognostic role of lymph node involvement.
- The model offers moderate predictive power (C-index ~0.67), confirmed by bootstrapping and cross-validation.

### Practical Implication
Clinically, combining Lev+5FU post-surgery could lead to better outcomes, especially in patients with higher node counts or advanced tumor extent, although further prospective validation is advised.
