---
title: "Nonparametric comparison of 2 or more groups"
author : Yuting Wu
output: html_document
date: "2025-03-10"
---
#This R script performs Survival Analysis with a focus on nonparametric comparison of 2 or more groups


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install_if_missing <- function(...) {
  packages <- c(...)
  new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
  if(length(new_packages)) install.packages(new_packages)
}


install_if_missing("ggplot2", "dplyr", "survival", "survminer", "tidyverse", "mice", "naniar", "VIM")


library(ggplot2)
library(dplyr)
library(survival)
library(survminer)
library(tidyverse)

library(naniar)
library(VIM)

```
#1. Load and process the data, check missing values, remove rows with missing values 

```{r data-prep}

colon <- read.csv("C:/Users/ywu09/Downloads/dataset-42025.csv")
summary(colon)
miss_var_summary(colon)
vis_miss(colon)
aggr(colon, prop = FALSE, numbers = TRUE)
colon_death <- subset(colon, etype == 2)
sum(is.na(colon_death))

# Remove rows with any missing values
colon_complete <- na.omit(colon)
sum(is.na(colon_complete))

```

#2. Nonparametric comparison of 2 or more groups_ treatment 


```{r kaplan-meier}
# Create a survival object 
surv_obj <- Surv(time = colon$time, event = colon$status)


fit_rx <- survfit(surv_obj ~ rx, data = colon)


smaller_theme <- theme(
  plot.title = element_text(size = 10),
  axis.title = element_text(size = 8),
  axis.text = element_text(size = 7),
  legend.title = element_text(size = 8),
  legend.text = element_text(size = 7),
  legend.position = "top"
)


km_plot <- ggsurvplot(fit_rx, 
                     data = colon,
                     pval = TRUE,           
                     risk.table = TRUE,     
                     conf.int = TRUE,       
                     xlab = "Time in days", 
                     legend.title = "Treatment",
                     legend.labs = c("Observation", "Levamisole", "Lev+5FU"),
                     surv.median.line = "hv",
                    
                     ggtheme = theme_bw() + smaller_theme,
                    
                     tables.theme = theme_cleantable() + smaller_theme,
                     risk.table.y.text.col = TRUE,
                     risk.table.y.text = FALSE,
                     risk.table.height = 0.25,
                    
                     pval.size = 3,
                     pval.coord = c(0, 0.1)  
)


km_plot$table <- km_plot$table + theme(
  axis.text = element_text(size = 6),
  axis.title = element_text(size = 7)
)


print(km_plot)
```
# Result: Significant difference in survival between treatment groups
##Lev+5FU vs. Observation: Significant survival benefit (adjusted p = 0.009).

#conclusion:




##Log-Rank Test and Pairwise
```{r Compares survival curves across treatment groups}
# Perform log-rank test
log_rank_test <- survdiff(surv_obj ~ rx, data = colon)
print(log_rank_test)



# Create a function for pairwise log-rank tests
pairwise_logrank <- function(formula, data, ...) {
  groups <- unique(eval(formula[[3]], data))
  combinations <- combn(groups, 2)
  results <- data.frame(group1 = character(), 
                        group2 = character(), 
                        p.value = numeric(),
                        stringsAsFactors = FALSE)
  
  for(i in 1:ncol(combinations)) {
    g1 <- combinations[1, i]
    g2 <- combinations[2, i]
    subset_data <- data[eval(formula[[3]], data) %in% c(g1, g2), ]
    test <- survdiff(formula, data = subset_data, ...)
    p.val <- 1 - pchisq(test$chisq, length(test$n) - 1)
    results <- rbind(results, data.frame(group1 = g1, group2 = g2, p.value = p.val))
  }
  
  return(results)
}

# Perform pairwise comparisons

# Create a survival object
surv_obj <- Surv(time = colon$time, event = colon$status)

# Perform pairwise comparisons
pairwise_results <- pairwise_logrank(Surv(time, status) ~ rx, data = colon)
print(pairwise_results)

# Apply Bonferroni correction for multiple testing
pairwise_results$p.adjusted <- p.adjust(pairwise_results$p.value, method = "bonferroni")
print(pairwise_results)

```


#Nonparametric comparison of 2 or more group___age
```{r age group}

colon_complete <- colon_complete |> 
  mutate(
    age_group = cut(age,
      breaks = c(0, 30, 60, 100),
      labels = c("<30", "30-60", ">60")
    )
)

# Verify group distribution
table(colon_complete$age_group)


# Log-Rank Test for age groups
survdiff(Surv(time, status) ~ age_group, data = colon_complete) |> 
  print()

# Kaplan-Meier Plot
ggsurvplot(
  survfit(Surv(time, status) ~ age_group, data = colon_complete),
  pval = TRUE, risk.table = TRUE, palette = "Dark2",
  title = "Age Group Comparison"
)
```
#conclusion: age significantly impacts recurrence-free survival (log-rank p=0.002). Patients <30 years old demonstrate the best outcomes (median survival 3021 days), while those >60 years have the poorest prognosis (median 1622 days). These differences persist after pairwise comparisons (p<0.05). 
#Cox model 
```{r Cox model}
cox_model <- coxph(
  Surv(time, status) ~ rx + sex + age + obstruct + nodes + differ + extent,
  data = colon_complete
)
summary(cox_model)


```


#check assumption 
```{r Cox model}
test_ph <- cox.zph(cox_model)
print(test_ph)  
ggcoxzph(test_ph)

```

