---
title: "Survival Analysis of Colon Cancer"
author: "Denis SOH"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Introduction

This analysis uses the `colon` dataset from the `{survival}` package to evaluate factors influencing survival time among colon cancer patients. We perform Kaplan-Meier survival estimation, log-rank tests, and Cox proportional hazards regression.

# Data Preparation

```{r setup, message=FALSE, warning=FALSE}
library(survival)
library(survminer)
library(dplyr)
library(broom)
library(ggplot2)
library(kableExtra)

# Load dataset
data(colon, package = "survival")
colon <- na.omit(colon)
colon <- colon[colon$etype == 2, ]

# Convert relevant variables to factors
colon$sex <- factor(colon$sex, labels = c("Male", "Female"))
colon$rx <- factor(colon$rx, levels = c("Obs", "Lev", "Lev+5FU"))
colon$obstruct <- factor(colon$obstruct, labels = c("No", "Yes"))

# Summary statistics
summary(colon[, c("time", "status", "age", "sex", "rx", "obstruct", "nodes")])
```

# Kaplan-Meier Survival Estimation

```{r km-overall, fig.width=7, fig.height=5}
library(survminer)
km_fit <- survfit(Surv(time, status) ~ rx, data = colon)
ggsurvplot(km_fit, data = colon, risk.table = TRUE, pval = TRUE,
           ggtheme = theme_minimal(),
           title = "Kaplan-Meier Survival Curves by Treatment")
```

The Kaplan-Meier curves indicate significant survival differences among treatment groups, with **Levamisole+5FU** showing better survival outcomes.

# Log-rank Test

```{r log-rank-test}
# Log-rank test by treatment
test_rx <- survdiff(Surv(time, status) ~ rx, data = colon)
test_result <- broom::tidy(test_rx)
kable(test_result, caption = "Log-rank test for survival differences among treatments") %>%
  kable_styling()
```

The Log-rank test confirms significant differences in survival between groups (`p < 0.001`).

# Cox Proportional Hazards Model

```{r cox-model}
cox_model <- coxph(Surv(time, status) ~ age + sex + obstruct + nodes + rx, data = colon)
summary(cox_model)

# HR Table
cox_results <- tidy(cox_model, exponentiate = TRUE, conf.int = TRUE)
cox_results <- cox_results %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

colnames(cox_results) <- c("Variable", "HR", "Lower CI", "Upper CI", "p-value")

cox_results %>%
  kable(caption = "Cox Regression Model Hazard Ratios") %>%
  kable_styling()
```

## Interpretation of Hazard Ratios

```{r}

# Installer les packages si nécessaire
if (!requireNamespace("broom", quietly = TRUE)) install.packages("broom")
if (!requireNamespace("kableExtra", quietly = TRUE)) install.packages("kableExtra")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")

library(broom)
library(kableExtra)
library(dplyr)

# Ajustement du modèle de Cox
cox_model <- coxph(Surv(time, status) ~ age + sex + obstruct + nodes + rx, data=colon)

# Extraction des résultats sous forme lisible avec Hazard Ratios
cox_results <- broom::tidy(cox_model, exponentiate = TRUE, conf.int = TRUE) %>% 
  dplyr::select(term, estimate, conf.low, conf.high, p.value) %>%
  dplyr::rename(
    Variable = term,
    HR = estimate,
    LowerCI = conf.low,
    UpperCI = conf.high,
    p = p.value
  ) %>%
  mutate(across(where(is.numeric), round, 3))

# Tableau avec interprétations personnalisées
cox_results$Interpretation <- c(
  "Small but significant increase in hazard per additional year of age.",
  "Not significant, no difference between females and males.",
  "Increased hazard if obstruction is present, not statistically significant after adjusting.",
  "Significant increase in hazard per additional positive lymph node.",
  "No significant difference compared to observation.",
  "Significant improvement in survival compared to observation."
)

# Afficher le tableau formaté
cox_results %>% 
  kableExtra::kable(caption = "Interpretation of Hazard Ratios from Cox Regression Model", align="lcccc") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

# Checking Proportional Hazards Assumption

```{r cox-ph}
zph <- cox.zph(cox_model)
ggcoxzph(zph)
```

The assumption of proportional hazards is reasonably met as no significant p-values appear.

# Model Performance: Harrell's C-index

```{r c-index}
cox_summary <- summary(cox_model)
cox_summary$concordance
```

The concordance index (Harrell's C-index) is around `r round(cox_summary$concordance[1], 2)`, indicating good predictive performance.

# Forest Plot of Hazard Ratios

```{r forest-plot, fig.width=7, fig.height=5}
ggforest(cox_model, data = colon)
```

The forest plot clearly highlights variables significantly impacting survival, notably the combination treatment (Lev+5FU) and the number of positive lymph nodes.

# Conclusion

This survival analysis highlights that treatment with **Levamisole+5FU** and **fewer positive nodes** significantly improve survival in colon cancer patients. Age slightly impacts survival, whereas sex and obstruction show minimal effects after adjustment. The Cox model is robust with adequate predictive performance (C-index ~0.7). Future analyses could investigate additional variables and interactions to refine these findings further.
